Modeling Intervention Strategies for United States TB Control

### 2013-10-13 in_progress/presentations/paper/draft.tex
added sensitivity analysis section to paper, updated basic reproduction number section

Further implementations of the \texttt{C++} model were made that deviated from
the basic Hill Model. Specifically, the acute latent compartment in the basic
Hill Model is a vestigal necessity of compartmental modeling and not reflective
of the biology of tuberculosis, and an agent based model was implemented that
more closely respects the biology of tuberculosis. Results from this model or
the strict Hill model did not differ significantly. 
\section{Results}
\subsection{Basic Population Breakdown}
The additional tracking capabilities offered several key insights into US TB
dynamics. Figure~\ref{fig:incPlotSourced} shows the yearly incidence of US TB, broken
down into infection source. It can be seen that the majority of the US TB load
is driven by activations of FB LTBI, followed by USB LTBI activations. This data
further agrees with the conclusions drawn by Hill, Becerra, and Castro about the
necessity of LTBI treatment in any valuable intervention strategy. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.5]{incPlotSourced}
  \end{center}
  \caption{Sourced yearly incidence data generated by the extended Hill Model.}
  \label{fig:incPlotSourced}
\end{figure}
Further, figure~\ref{fig:costPlotSourced} shows a similarly sourced plot, but analyzing the
final US HCS costs due to TB. One can see that in this plot, roughly half of the
US TB HCS costs are due to activations of LTBI. In this data, it is appropriate
to think of the costs due to acute LTBI are due to exogenous re-inection,
whereas active TB costs due to chronic LTBI are activations of LTBI into active
TB. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.35]{costPlotSourced}
  \end{center}
  \caption{Sourced US HCS economic TB load. Note that this data only illustrates
    the load due to treating active TB, but illustrates where the infections
    driving this cost come from.}
  \label{fig:costPlotSourced}
\end{figure}
Note that both of these plots underestimate the impact LTBI activations play in
the spread of TB, as every LTBI activation to infectious TB contributes not only
to incidence and costs directly, but also indirectly by causing additional
future cases, which is not captured in these graphs. Further, there is also US
HCS costs due to treating LTBI, which is not illustrated in these graphs. 
\subsubsection{Basic Reproduction Number}
The basic reproduction number of FB or USB cases of infectious TB was also
estimated by this system. From a theoretical perspective, we can think of the
total number of secondary infections over 100 years due to a FB or USB
infectious TB case as describing a geometric series in a large population.
Presuming there are no overlaps in infectious contacts, if a single case of
infectious TB in either population infects $p_f, p_u$ new cases in one year,
respectively, then we can say that over the course of 100 years, the total
number of cases infected will follow a geometric series. This analysis predicts
that over 100 years, one USB infection will lead to $1.03$ subsequent infections,
whereas one FB infection will lead to $.64$ subsequent infections. The full calculations
are included in the appendix.  Experimentally, these data were also analyzed, 
with results of $1.04$ and $.8$, respectively. WHAT CAUSES THIS DIFFERENCE? WE
NEED TO ANSWER THAT.

@@

\subsection{Sensitivity Analysis}
\section{Discussion}
These results confirm the hypothesis that curing incoming LTBI rates is a
necessary step towards elimination and indicate that it is a cost effective
option. 
\subsection{Future Work}
This work could be extended by examining different classes of interventions or
more accurately estimating intervention cost with the deterministic extended
Hill. Further work could also be done with the agent-based Hill Model, by using
it to examine the effect contact structure plays on US TB incidence levels or to
examine the effects drug-resistant TB will have on US TB dynamics. 
\section{Appendix}
\subsection{Hill Constants}
Below, we detail some of the relevant constants in the Basic Hill Model. A full
listing of constants used in the original Hill Model can be found in CITATION!
\begin{figure}[h]
  \begin{center}
    \tiny{
    \begin{verbatim}
sigmaLBase  <- 0.057
fBase       <- 0.187
transBase   <- 1
incLTBIBase <- 1

@@

#Cost calculations
dcL0 <- discV * Cl * sigmaL  * 1e6 * L0                    #cost for Chronic LTBI cures      (USB)
dcF0 <- discV * Cl * sigmaF0 * 1e6 * F0                    #cost for Acute LTBI cures        (USB)
dcI0 <- discV * Ct * q*(dprogAcute0 + dprogChron0) * 1e6     #cost for Infectious TB cures     (USB)
dcJ0 <- discV * Ct * (1-q)*(dprogAcute0 + dprogChron0) * 1e6 #cost for Non-Infectious TB cures (USB)
dcI0dL0 <- discV * Ct * q*(dprogChron0) * 1e6     #cost for Infectious TB cures     (USB)
dcI0dF0 <- discV * Ct * q*(dprogChron0) * 1e6     #cost for Infectious TB cures     (USB)
dcJ0dL0 <- discV * Ct * (1-q)*(dprogAcute0) * 1e6 #cost for Non-Infectious TB cures (USB)
dcJ0dF0 <- discV * Ct * (1-q)*(dprogAcute0) * 1e6 #cost for Non-Infectious TB cures (USB)
dcL1 <- discV * Cl * sigmaL  * 1e6 * L1                    #cost for Chronic LTBI cures      (FB)
dcF1 <- discV * Cl * sigmaF1 * 1e6 * F1                    #cost for Acute LTBI cures        (FB)
dcI1 <- discV * Ct * q*(dprogAcute1 + dprogChron1) * 1e6     #cost for Infectious TB cures     (FB)
dcJ1 <- discV * Ct * (1-q)*(dprogAcute1 + dprogChron1) * 1e6 #cost for Non-Infectious TB cures (FB)
dcI1dL1 <- discV * Ct * q*(dprogChron1) * 1e6     #cost for Infectious TB cures     (USB)
dcI1dF1 <- discV * Ct * q*(dprogChron1) * 1e6     #cost for Infectious TB cures     (USB)
dcJ1dL1 <- discV * Ct * (1-q)*(dprogAcute1) * 1e6 #cost for Non-Infectious TB cures (USB)
dcJ1dF1 <- discV * Ct * (1-q)*(dprogAcute1) * 1e6 #cost for Non-Infectious TB cures (USB)
dcN0 <- dcF0 + dcL0 + dcI0 + dcJ0                            #Total cost for all treatments (USB)
dcN1 <- dcF1 + dcL1 + dcI1 + dcJ1                            #Total cost for all treatments (FB)
    \end{verbatim}
  }
  \end{center}
  \label{fig:extendedHillEqs}
  \caption{
    The equations defining the Extended Hill Model, shown in \texttt{R} syntax.
    In this model, compartments share the same names as they did in the Hill
    Model, \emph{i.e.} \texttt{S1} is the Foreign Born susceptible population. 
  }
\end{figure}
Here, a continuous discounting approximation is used to a discount rate of $3\%$
yearly, found via $discV$. In many of the differential equations, a small $c$ is
used to denote \emph{cost of}. 
\subsection{Estimations for Infectious Rates of FB and USB}
Colin to add. 
\subsection{Statistical Qualities of the Agent Based Model}
\subsection{Finer Intervention Analysis}
\subsection{Sensitivity Analysis}
\end{document}

>>>>

Further implementations of the \texttt{C++} model were made that deviated from
the basic Hill Model. Specifically, the acute latent compartment in the basic
Hill Model is a vestigal necessity of compartmental modeling and not reflective
of the biology of tuberculosis, and an agent based model was implemented that
more closely respects the biology of tuberculosis. Results from this model or
the strict Hill model did not differ significantly. 
\section{Results}
\subsection{Basic Population Breakdown}
The additional tracking capabilities offered several key insights into US TB
dynamics. Figure~\ref{fig:incPlotSourced} shows the yearly incidence of US TB, broken
down into infection source. It can be seen that the majority of the US TB load
is driven by activations of FB LTBI, followed by USB LTBI activations. This data
further agrees with the conclusions drawn by Hill, Becerra, and Castro about the
necessity of LTBI treatment in any valuable intervention strategy. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.5]{incPlotSourced}
  \end{center}
  \caption{Sourced yearly incidence data generated by the extended Hill Model.}
  \label{fig:incPlotSourced}
\end{figure}
Further, figure~\ref{fig:costPlotSourced} shows a similarly sourced plot, but analyzing the
final US HCS costs due to TB. One can see that in this plot, roughly half of the
US TB HCS costs are due to activations of LTBI. In this data, it is appropriate
to think of the costs due to acute LTBI are due to exogenous re-inection,
whereas active TB costs due to chronic LTBI are activations of LTBI into active
TB. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.35]{costPlotSourced}
  \end{center}
  \caption{Sourced US HCS economic TB load. Note that this data only illustrates
    the load due to treating active TB, but illustrates where the infections
    driving this cost come from.}
  \label{fig:costPlotSourced}
\end{figure}
Note that both of these plots underestimate the impact LTBI activations play in
the spread of TB, as every LTBI activation to infectious TB contributes not only
to incidence and costs directly, but also indirectly by causing additional
future cases, which is not captured in these graphs. Further, there is also US
HCS costs due to treating LTBI, which is not illustrated in these graphs. 
\subsubsection{Basic Reproduction Number}
The basic reproduction number of FB or USB cases of infectious TB was also
estimated by this system. Using the best-fit values of parameters from the 
Hill model, we found the average number of secondary infections arising from
a single case of infectious TB to be 0.423 and 0.370 for USB and FB populations
respectively.  These values differ because in the model, there are different
estimates for death rates and the rate of TB treatment and self-cure in the USB and FB populations, 
so the average infectious period differs between the two populations.
Extrapolating these results, we can think of the
total number of secondary infections over 100 years due to a FB or USB
infectious TB case as describing a geometric series in a large population.
Presuming there are no overlaps in infectious contacts, if a single case of
infectious TB in either population infects $p_f, p_u$ new cases,
respectively, then we can say that over the course of 100 years, the total
number of cases infected will follow a geometric series. This analysis predicts
that over 100 years, one USB infection will lead to $1.04$ subsequent infections,
whereas one FB infection will lead to $.83$ subsequent infections. Experimentally, 
these data were also analyzed, with results of $1.03$ and $.64$, respectively. 
Full calculations are included in the appendix.

@@

\subsection{Sensitivity Analysis}
To account for uncertainty of input parameter values and to gain insight about the
most influential parameters in the model, we used Latin Hypercube Sampling to 
analyze the deterministic model, implemented with the 'lhs' package in R.  We varied 
the 16 parameters from the original Hill model according to triangular distributions centered
around their best-fit values in the Hill model, and two additional treatment cost parameters according to uniform distributions.  We obtained similar results to the Hill model 
analyzing non-economic outcomes, as expected for validation.  For economic 
outcomes, we found the parameter $f$, the fraction of FB arrivals with LTBI,
to be highly correlated with the projected overall cost burden of TB in the United States
over the next 100 years.  Another highly influential parameter in the model was $\sigma^{L}$, 
the treatment rate for chronic LTBI.  However, while increasing $\sigma^{L}$ significantly
decreases the cost burden due to Active TB treatment, it increases the cost burden due to Latent TB treatment by a greater amount, given the model estimates that Latent TB treament and Active
TB treatment health care costs are \$14,014.90 \$403.45 per case cured respectively.  
A full description of the Latin Hypercube Sampling analysis and tables of Partial Rank Correlation Coefficient (PRCC) results for all parameters are given in the appendix, section X.

\section{Discussion}
These results confirm the hypothesis that curing incoming LTBI rates is a
necessary step towards elimination and indicate that it is a cost effective
option. 
\subsection{Future Work}
This work could be extended by examining different classes of interventions or
more accurately estimating intervention cost with the deterministic extended
Hill. Further work could also be done with the agent-based Hill Model, by using
it to examine the effect contact structure plays on US TB incidence levels or to
examine the effects drug-resistant TB will have on US TB dynamics. 
\section{Appendix}
\subsection{Hill Constants}
Below, we detail some of the relevant constants in the Basic Hill Model. A full
listing of constants used in the original Hill Model can be found in CITATION!
\begin{figure}[h]
  \begin{center}
    \tiny{
    \begin{verbatim}
sigmaLBase  <- 0.057
fBase       <- 0.187
transBase   <- 1
incLTBIBase <- 1

@@

#Cost calculations
dcL0 <- discV * Cl * sigmaL  * 1e6 * L0                    #cost for Chronic LTBI cures      (USB)
dcF0 <- discV * Cl * sigmaF0 * 1e6 * F0                    #cost for Acute LTBI cures        (USB)
dcI0 <- discV * Ct * q*(dprogAcute0 + dprogChron0) * 1e6     #cost for Infectious TB cures     (USB)
dcJ0 <- discV * Ct * (1-q)*(dprogAcute0 + dprogChron0) * 1e6 #cost for Non-Infectious TB cures (USB)
dcI0dL0 <- discV * Ct * q*(dprogChron0) * 1e6     #cost for Infectious TB cures     (USB)
dcI0dF0 <- discV * Ct * q*(dprogChron0) * 1e6     #cost for Infectious TB cures     (USB)
dcJ0dL0 <- discV * Ct * (1-q)*(dprogAcute0) * 1e6 #cost for Non-Infectious TB cures (USB)
dcJ0dF0 <- discV * Ct * (1-q)*(dprogAcute0) * 1e6 #cost for Non-Infectious TB cures (USB)
dcL1 <- discV * Cl * sigmaL  * 1e6 * L1                    #cost for Chronic LTBI cures      (FB)
dcF1 <- discV * Cl * sigmaF1 * 1e6 * F1                    #cost for Acute LTBI cures        (FB)
dcI1 <- discV * Ct * q*(dprogAcute1 + dprogChron1) * 1e6     #cost for Infectious TB cures     (FB)
dcJ1 <- discV * Ct * (1-q)*(dprogAcute1 + dprogChron1) * 1e6 #cost for Non-Infectious TB cures (FB)
dcI1dL1 <- discV * Ct * q*(dprogChron1) * 1e6     #cost for Infectious TB cures     (USB)
dcI1dF1 <- discV * Ct * q*(dprogChron1) * 1e6     #cost for Infectious TB cures     (USB)
dcJ1dL1 <- discV * Ct * (1-q)*(dprogAcute1) * 1e6 #cost for Non-Infectious TB cures (USB)
dcJ1dF1 <- discV * Ct * (1-q)*(dprogAcute1) * 1e6 #cost for Non-Infectious TB cures (USB)
dcN0 <- dcF0 + dcL0 + dcI0 + dcJ0                            #Total cost for all treatments (USB)
dcN1 <- dcF1 + dcL1 + dcI1 + dcJ1                            #Total cost for all treatments (FB)
    \end{verbatim}
  }
  \end{center}
  \label{fig:extendedHillEqs}
  \caption{
    The equations defining the Extended Hill Model, shown in \texttt{R} syntax.
    In this model, compartments share the same names as they did in the Hill
    Model, \emph{i.e.} \texttt{S1} is the Foreign Born susceptible population. 
  }
\end{figure}
Here, a continuous discounting approximation is used to a discount rate of $3\%$
yearly, found via $discV$. In many of the differential equations, a small $c$ is
used to denote \emph{cost of}. 
\subsection{Estimations for Infectious Rates of FB and USB}

\subsection{Statistical Qualities of the Agent Based Model}
\subsection{Finer Intervention Analysis}
\subsection{Sensitivity Analysis}
\end{document}
